---
title: "A1 - Getting started with ALprekDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A1 - Getting started with ALprekDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ALprekDB is a public R package for private Alabama First Class Pre-K
administrative data workflows. It turns annual ADECE budget, classroom, and
student files into validated panels, linked master datasets, and optional
DuckDB outputs.

Public examples in the package use synthetic data. Real ADECE files and
row-level outputs should stay in private local projects.

The v0.6.0 real-data scope is asymmetric:

| Module | Covered school years | Notes |
|--------|----------------------|-------|
| Budget | 2021-22 through 2024-25 | 2025-26 budget data are not inferred or zero-filled. |
| Classroom | 2021-22 through 2025-26 | Classroom codes use six-digit program codes. |
| Student | 2021-22 through 2025-26 | Student PII is excluded by default in private workflows. |

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("joonho112/ALprekDB")
```

```{r load}
library(ALprekDB)
```

## Classroom Codes

Budget, classroom, and student records are linked by classroom code. The
current contract is `CCCDNNNNNN.NN`:

- `CCC` is a three-digit county code;
- `D` is the delivery type code;
- `NNNNNN` is a six-digit program code;
- `NN` is the classroom number within site.

The examples below use fake `9xx` prefixes and `9xxxxx` program codes for
public documentation.

```{r parse-classroom-codes}
codes <- c("901P900001.01", "967H900002.02", "933C900003.01")
parse_classroom_codes(codes)
```

Delivery type labels are data-driven:

```{r delivery-types}
alprek_delivery_types()
```

## Codebooks and Mappings

ALprekDB uses CSV codebooks stored in `inst/extdata/` for column mappings,
budget categories, delivery types, race/ethnicity mappings, and other
standardization rules. The helpers expose those mappings as ordinary tibbles.

```{r category-groups}
head(alprek_category_groups(), 10)
```

## Synthetic Quick Start

The synthetic generators create panel objects directly. When the same
`n_classrooms`, `n_years`, and seed are used, the generated budget, classroom,
and student panels share classroom codes that can be linked. Public synthetic
outputs intentionally use fake `9xx` classroom-code prefixes and synthetic
county labels so printed rows cannot be mistaken for confidential ADECE
records.

```{r synthetic-panels}
budget <- alprek_synthetic_budget(
  n_classrooms = 20,
  n_years = 2,
  seed = 42
)

classroom <- alprek_synthetic_classroom(
  n_classrooms = 20,
  n_years = 2,
  seed = 42
)

student <- alprek_synthetic_student(
  n_students = 100,
  n_classrooms = 20,
  n_years = 2,
  seed = 42
)

budget
classroom
student
```

A quick look at the classroom panel shows the shared keys that make linkage
possible:

```{r inspect-classroom-panel}
head(classroom$data[c(
  "school_year",
  "classroom_code",
  "county_code",
  "delivery_type",
  "site_name"
)])
```

## Link the Modules

`linkage_create_master()` creates a two-level master object:

- `classroom_level`: one row per classroom-year, with budget fields and
  classroom-level student aggregates;
- `student_level`: one row per student-year, with classroom and budget
  attributes attached.

```{r create-master}
master <- linkage_create_master(
  budget = budget,
  classroom = classroom,
  student = student
)

master
```

The linkage object carries validation and coverage diagnostics:

```{r validate-master}
linkage_validate(master)
```

```{r linkage-summary}
linkage_summary_stats(master)
```

```{r coverage-summary}
master$meta$coverage$by_year[c(
  "school_year",
  "has_budget",
  "has_classroom",
  "has_student",
  "n_budget_rows",
  "n_classroom_rows",
  "n_student_rows"
)]
```

The classroom-level master is the usual starting point for classroom or
site-level descriptive analysis:

```{r classroom-master-preview}
head(master$classroom_level[c(
  "school_year",
  "classroom_code",
  "delivery_type",
  "n_children",
  "grand_total",
  "per_child_budget",
  "per_seat_budget"
)])
```

## First Private Real-Data Recipe

Real-data examples should be run only in a private project. Keep paths in
environment variables, keep row-level outputs out of GitHub, and keep student
PII/date-of-birth fields excluded unless a controlled analysis explicitly
requires them.

```{r real-data-config, eval = FALSE}
data_dir <- Sys.getenv("ALPREKDB_DATA_DIR")

budget_cfg <- budget_config(
  school_year = "2024-2025",
  budget_path = file.path(data_dir, "budget_2024_2025.xlsx")
)

classroom_cfg <- classroom_config(
  school_year = "2024-2025",
  classroom_path = file.path(data_dir, "classroom_2024_2025.xlsx"),
  include_dob = FALSE
)

student_cfg <- student_config(
  school_year = "2024-2025",
  path = file.path(data_dir, "student_2024_2025.xlsx"),
  include_pii = FALSE
)
```

Process one module first, inspect its validation object, and then expand to the
full multi-module workflow:

```{r real-data-process-one-year, eval = FALSE}
budget_result <- budget_process(budget_cfg, export = FALSE)
budget_result$validation
budget_result$long
budget_result$master

classroom_result <- classroom_process(classroom_cfg, export = FALSE)
classroom_result$validation

student_result <- student_process(student_cfg, export = FALSE)
student_result$validation
```

For linked private analysis, use the multi-year processing helpers even when
starting with a single year. That keeps the object classes consistent with
larger panel workflows.

```{r real-data-link-one-year, eval = FALSE}
budget_panel <- budget_process_years(list(budget_cfg), export = FALSE)$panel
classroom_panel <- classroom_process_years(list(classroom_cfg), export = FALSE)$panel
student_panel <- student_process_years(list(student_cfg), export = FALSE)$panel

master <- linkage_create_master(
  budget = budget_panel,
  classroom = classroom_panel,
  student = student_panel
)

linkage_validate(master)
linkage_summary_stats(master)
```

## Private Targets Workflow

For repeated real-data processing, start from the package's `targets` template
in a private project directory:

```{r copy-targets-template, eval = FALSE}
template_dir <- system.file("templates", "targets", package = "ALprekDB")

dir.create("alprekdb-private-workflow", showWarnings = FALSE)
file.copy(
  list.files(template_dir, full.names = TRUE, all.files = TRUE, no.. = TRUE),
  "alprekdb-private-workflow",
  recursive = TRUE
)
```

Then configure the private workflow outside the package repository:

```sh
export ALPREKDB_RUN_REALDATA=1
export ALPREKDB_DATA_DIR="/path/to/local/ADECE/source/files"
export ALPREKDB_OUTPUT_DIR="output/alprekdb"
```

Row-level RDS outputs and DuckDB writes require an explicit local opt-in:

```sh
export ALPREKDB_WRITE_OUTPUTS=1
```

## Where to Go Next

- A2 - Build panels: budget, classroom, and student panel construction.
- A3 - Linkage analysis: orphan diagnostics and derived measures.
- A4 - DuckDB and SQL: persistent local database workflows.
- A5 - Targets workflow: reproducible private processing.
- M4 - Privacy and provenance: guardrails for private source data and outputs.
