---
title: "A2 - Build Panels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A2 - Build Panels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ALprekDB builds three panel objects before linkage:

- a budget panel with one row per classroom-year funding record;
- a classroom panel with one row per classroom-year site/classroom record;
- a student panel with one row per student-year record.

In v0.6.0, the real-data source scope is intentionally asymmetric:

| Module | Covered school years | Panel policy |
|--------|----------------------|--------------|
| Budget | 2021-22 through 2024-25 | Do not infer, copy forward, or zero-fill 2025-26 budget records. |
| Classroom | 2021-22 through 2025-26 | Retain classroom rows even when budget is unavailable. |
| Student | 2021-22 through 2025-26 | Exclude PII by default and retain student rows even when budget is unavailable. |

This vignette uses synthetic data for executable examples. The private
real-data snippets show the same object contract without exposing raw ADECE
files.

```{r load}
library(ALprekDB)
```

## Build Synthetic Panels

The synthetic generators return panel objects directly. They are useful for
learning the package structure and testing analysis code, not for estimating
Alabama program quantities. Generated classroom identifiers use fake `9xx`
prefixes and synthetic county labels to keep public printed output visibly
separate from confidential ADECE rows.

```{r synthetic-panels}
n_classrooms <- 12
n_students <- 60
seed <- 42

budget_panel <- alprek_synthetic_budget(
  n_classrooms = n_classrooms,
  n_years = 2,
  seed = seed
)

classroom_panel <- alprek_synthetic_classroom(
  n_classrooms = n_classrooms,
  n_years = 3,
  seed = seed
)

student_panel <- alprek_synthetic_student(
  n_students = n_students,
  n_classrooms = n_classrooms,
  n_years = 3,
  seed = seed
)

budget_panel
classroom_panel
student_panel
```

This synthetic setup deliberately gives budget two years and classroom/student
three years. It mirrors the real v0.6.0 lesson: a later classroom/student year
can be retained even when budget coverage is unavailable.

## Panel Object Anatomy

Every panel object has a `$data` table and a `$years` vector. Module-specific
metadata differs slightly.

```{r panel-anatomy}
panel_anatomy <- function(x, module) {
  data.frame(
    module = module,
    class = paste(class(x), collapse = ", "),
    fields = paste(names(x), collapse = ", "),
    years = paste(as.character(x$years), collapse = ", "),
    rows = nrow(x$data),
    columns = ncol(x$data),
    stringsAsFactors = FALSE
  )
}

rbind(
  panel_anatomy(budget_panel, "budget"),
  panel_anatomy(classroom_panel, "classroom"),
  panel_anatomy(student_panel, "student")
)
```

The main difference is that budget panels expose `$n_years`, while classroom
and student panels expose `$n_total`. Classroom panels also carry an
`$imputation_log`; student panels carry `$n_unique_students`.

## Budget Panel

Budget panels track OSR and other funding sources by classroom-year.

```{r budget-preview}
head(budget_panel$data[c(
  "school_year",
  "classroom_code",
  "delivery_type",
  "grand_total",
  "share_osr"
)])
```

```{r budget-summary}
budget_summary_stats(budget_panel)[c(
  "school_year",
  "n",
  "grand_total_mean",
  "grand_total_median",
  "share_osr_mean"
)]
```

```{r budget-tracking}
head(budget_track_classrooms(budget_panel))
```

## Classroom Panel

Classroom panels hold site, geography, staffing, delivery type, and classroom
metadata.

```{r classroom-preview}
head(classroom_panel$data[c(
  "school_year",
  "classroom_code",
  "county_code",
  "delivery_type",
  "site_name",
  "latitude",
  "longitude"
)])
```

```{r classroom-summary}
classroom_summary_stats(classroom_panel)
```

The imputation log records forward-filled coordinates and first-funded years
within site groups when those operations are needed.

```{r classroom-imputation}
classroom_panel$imputation_log
```

## Student Panel

Student panels contain one row per student-year. Direct PII is excluded by
default in private real-data workflows; synthetic examples use fabricated
student identifiers.

```{r student-preview}
head(student_panel$data[c(
  "school_year",
  "classroom_code",
  "adece_id",
  "gender",
  "race",
  "poverty_dum",
  "days_absent_total"
)])
```

```{r student-summary}
student_summary_stats(student_panel)
```

`student_transform()` adds derived analysis variables such as assessment gains,
chronic absence, service density, and risk index when the source columns are
available.

```{r student-transform}
enriched_student_panel <- student_transform(student_panel)

head(enriched_student_panel$data[c(
  "school_year",
  "classroom_code",
  "gold_literacy_gain_raw",
  "chronic_absence",
  "n_services",
  "risk_index"
)])
```

```{r student-transform-log}
enriched_student_panel$transform_log
```

## Coverage Caveat Before Linkage

The panel builder should preserve module-specific coverage. Later linkage steps
make the coverage gap explicit rather than fabricating budget records.

```{r synthetic-coverage-linkage}
coverage_master <- linkage_create_master(
  budget = budget_panel,
  classroom = classroom_panel,
  student = enriched_student_panel
)

coverage_master$meta$coverage$by_year[c(
  "school_year",
  "has_budget",
  "has_classroom",
  "has_student",
  "n_budget_rows",
  "n_classroom_rows",
  "n_student_rows",
  "budget_status"
)]
```

```{r synthetic-coverage-validation}
linkage_validate(coverage_master)
```

In real v0.6.0 processing, this same pattern applies to 2025-26: classroom and
student rows are retained, and budget-derived fields remain missing until a
canonical 2025-26 budget source exists.

## Private Real-Data Manifest

For private processing, define a local manifest from environment-variable
paths. Keep the manifest in a private project and keep `include_pii = FALSE`
and `include_dob = FALSE` unless a controlled analysis explicitly requires
otherwise.

```{r private-manifest, eval = FALSE}
data_dir <- Sys.getenv("ALPREKDB_DATA_DIR")

budget_files <- c(
  "2021-2022" = "budget_2021_2022.xlsx",
  "2022-2023" = "budget_2022_2023.xlsx",
  "2023-2024" = "budget_2023_2024.xlsx",
  "2024-2025" = "budget_2024_2025.xlsx"
)

classroom_files <- c(
  "2021-2022" = "classroom_2021_2022.xlsx",
  "2022-2023" = "classroom_2022_2023.xlsx",
  "2023-2024" = "classroom_2023_2024.xlsx",
  "2024-2025" = "classroom_2024_2025.xlsx",
  "2025-2026" = "classroom_2025_2026.xlsx"
)

student_files <- c(
  "2021-2022" = "student_2021_2022.xlsx",
  "2022-2023" = "student_2022_2023.xlsx",
  "2023-2024" = "student_2023_2024.xlsx",
  "2024-2025" = "student_2024_2025.xlsx",
  "2025-2026" = "student_2025_2026.xlsx"
)
```

Notice that there is no 2025-26 budget entry. That absence is an input contract,
not a missing coding step.

## Private Real-Data Processing

The `*_process_years()` helpers return a result object with a `$panel` element
and a `$validation_summary` element. These examples are not executed in public
documentation.

```{r private-configs, eval = FALSE}
budget_configs <- Map(
  function(year, file) {
    budget_config(
      school_year = year,
      budget_path = file.path(data_dir, file),
      verbose = FALSE
    )
  },
  names(budget_files),
  budget_files
)

classroom_configs <- Map(
  function(year, file) {
    classroom_config(
      school_year = year,
      classroom_path = file.path(data_dir, file),
      include_dob = FALSE
    )
  },
  names(classroom_files),
  classroom_files
)

student_configs <- Map(
  function(year, file) {
    student_config(
      school_year = year,
      path = file.path(data_dir, file),
      include_pii = FALSE
    )
  },
  names(student_files),
  student_files
)
```

```{r private-process-years, eval = FALSE}
budget_processed <- budget_process_years(budget_configs, export = FALSE)
classroom_processed <- classroom_process_years(classroom_configs, export = FALSE)
student_processed <- student_process_years(student_configs, export = FALSE)

budget_panel <- budget_processed$panel
classroom_panel <- classroom_processed$panel
student_panel <- student_transform(student_processed$panel)

budget_processed$validation_summary
classroom_processed$validation_summary
student_processed$validation_summary
```

Row-level exports, local DuckDB files, and `_targets/` caches should remain
outside Git and outside public package artifacts. Use `ALPREKDB_WRITE_OUTPUTS=1`
only in a private workflow when those outputs are intentionally needed.

## Output Contract

| Helper family | Main output contract |
|---------------|----------------------|
| `budget_process()` | `raw`, `long`, `validation`, and `master` |
| `classroom_process()` | `raw`, `clean`, and `validation` |
| `student_process()` | `raw`, `clean`, and `validation` |
| `*_process_years()` | `by_year`, `panel`, and `validation_summary` |
| Budget panels | `$data`, `$years`, `$n_years`, and `$by_year` |
| Classroom panels | `$data`, `$years`, `$n_total`, `$by_year`, and `$imputation_log` |
| Student panels | `$data`, `$years`, `$n_total`, `$n_unique_students`, and `$by_year` |

## Next Step

A3 links the three panel objects into classroom-level and student-level master
datasets, then reads the orphan and coverage diagnostics created during that
join.
