---
title: "A3 - Linkage Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A3 - Linkage Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ALprekDB links budget, classroom, and student panels into a two-level master
object:

- `classroom_level`: one row per classroom-year, with budget fields and
  classroom-level student aggregates;
- `student_level`: one row per student-year, with classroom and budget
  attributes attached.

The main linkage rule is simple: missing coverage years are metadata, while
overlap-year orphans are linkage problems to inspect. In v0.6.0, the real-data
workflow keeps 2025-26 classroom and student rows even though the canonical
2025-26 budget file is not available.

```{r load}
library(ALprekDB)
```

## Build Synthetic Panels

This public example intentionally gives budget fewer years than classroom and
student panels. The final synthetic year stands in for the v0.6.0 2025-26
budget caveat. The synthetic classroom identifiers use fake `9xx` prefixes so
row-level excerpts in this vignette remain visibly separate from confidential
ADECE records.

```{r synthetic-panels}
n_classrooms <- 12
n_students <- 60
seed <- 42

budget_panel <- alprek_synthetic_budget(
  n_classrooms = n_classrooms,
  n_years = 2,
  seed = seed
)

classroom_panel <- alprek_synthetic_classroom(
  n_classrooms = n_classrooms,
  n_years = 3,
  seed = seed
)

student_panel <- alprek_synthetic_student(
  n_students = n_students,
  n_classrooms = n_classrooms,
  n_years = 3,
  seed = seed
)

student_panel <- student_transform(student_panel)
```

## Direct Joins

The master helper performs all joins, but the direct join helpers are useful for
understanding diagnostics.

```{r classroom-budget-join}
classroom_budget <- linkage_classroom_budget(
  classroom = classroom_panel,
  budget = budget_panel
)

classroom_budget
```

```{r student-classroom-join}
student_classroom <- linkage_student_classroom(
  student = student_panel,
  classroom = classroom_panel
)

student_classroom
```

## Create the Master Object

`linkage_create_master()` builds the classroom-level and student-level analysis
tables and stores the diagnostics from both join paths.

```{r create-master}
master <- linkage_create_master(
  budget = budget_panel,
  classroom = classroom_panel,
  student = student_panel
)

master
```

```{r master-assertions, include = FALSE}
validation <- linkage_validate(master)
needed_coverage_cols <- c(
  "school_year",
  "has_budget",
  "has_classroom",
  "has_student",
  "budget_status"
)
coverage_by_year <- master$meta$coverage$by_year
if (is.null(coverage_by_year) ||
    !all(needed_coverage_cols %in% names(coverage_by_year))) {
  coverage_by_year <- master$diagnostics$coverage$by_year
}
coverage_by_year <- as.data.frame(coverage_by_year)
stopifnot(inherits(master, "alprek_linkage_master"))
stopifnot(is.data.frame(master$classroom_level))
stopifnot(is.data.frame(master$student_level))
stopifnot(all(needed_coverage_cols %in% names(coverage_by_year)))
stopifnot(all(c("budget_overlap_orphans", "budget_missing_coverage") %in%
  validation$checks$check_name))
stopifnot(!any(grepl("[.]x$|[.]y$", names(master$classroom_level))))
stopifnot(!any(grepl("[.]x$|[.]y$", names(master$student_level))))
```

## Coverage Diagnostics

Coverage is summarized by year. A missing budget year is retained as an
explicit coverage status; it is not turned into a zero budget.

```{r coverage-by-year}
coverage_by_year[c(
  "school_year",
  "has_budget",
  "has_classroom",
  "has_student",
  "n_budget_rows",
  "n_classroom_rows",
  "n_student_rows",
  "budget_status"
)]
```

`linkage_summary_stats()` keeps budget means as `NA` when a year has no budget
coverage.

```{r linkage-summary}
linkage_summary_stats(master)
```

## Orphan Diagnostics

The classroom-budget diagnostics distinguish overlap-year orphans from rows
retained because an entire budget year is unavailable.

```{r classroom-budget-orphans}
classroom_budget$diagnostics$orphan_summary_by_year
```

```{r classroom-budget-metrics}
paste_or_none <- function(x) {
  if (length(x) == 0) "none" else paste(x, collapse = ", ")
}

data.frame(
  metric = c(
    "classroom rows without budget in overlapping years",
    "classroom rows retained because budget year is missing",
    "budget rows without classroom in overlapping years",
    "missing budget years"
  ),
  value = c(
    classroom_budget$diagnostics$n_left_orphan_overlap_years,
    classroom_budget$diagnostics$n_left_orphan_missing_budget_years,
    classroom_budget$diagnostics$n_right_orphan_overlap_years,
    paste_or_none(classroom_budget$diagnostics$missing_budget_years)
  ),
  stringsAsFactors = FALSE
)
```

The student-classroom diagnostics show whether student classroom codes matched
classroom records and whether any classrooms had no linked student rows.

```{r student-classroom-orphans}
student_classroom$diagnostics$orphan_summary_by_year
```

```{r student-classroom-metrics}
data.frame(
  metric = c(
    "student classroom codes missing in overlapping years",
    "student rows affected in overlapping years",
    "classrooms with no linked student rows",
    "missing classroom years",
    "missing student years"
  ),
  value = c(
    student_classroom$diagnostics$n_student_orphan_overlap_years,
    student_classroom$diagnostics$n_student_orphan_overlap_year_rows,
    student_classroom$diagnostics$n_classroom_orphan,
    paste_or_none(student_classroom$diagnostics$missing_classroom_years),
    paste_or_none(student_classroom$diagnostics$missing_student_years)
  ),
  stringsAsFactors = FALSE
)
```

## Validation Checks

`linkage_validate()` reports pass, warning, error, and info checks. In this
example the missing budget year is informational because it is an explicit
coverage gap, not a broken classroom-budget key in an overlapping year.

```{r validation}
validation
```

```{r validation-table}
validation$checks[c(
  "check_name",
  "status",
  "n_issues",
  "details"
)]
```

## Student Aggregation

The classroom-level master includes student aggregates created by
`linkage_aggregate_students()`. You can also call the aggregation helper
directly.

```{r aggregate-students}
student_aggregates <- linkage_aggregate_students(student_panel)

head(student_aggregates[c(
  "school_year",
  "classroom_code",
  "n_children",
  "pct_poverty",
  "mean_age",
  "pct_chronic_absence",
  "mean_risk_index"
)])
```

## Classroom-Level Analysis

Budget-derived fields are available for years with budget coverage. They remain
missing for years without budget coverage.

```{r classroom-analysis-preview}
cl <- master$classroom_level

head(cl[c(
  "school_year",
  "classroom_code",
  "delivery_type",
  "n_children",
  "grand_total",
  "per_child_budget",
  "per_seat_budget",
  "pct_poverty",
  "mean_days_absent"
)])
```

```{r classroom-analysis-summary}
available_budget <- !is.na(cl$per_child_budget)

aggregate(
  cbind(per_child_budget, pct_poverty, mean_days_absent) ~ school_year,
  data = cl[available_budget, ],
  FUN = function(x) round(mean(x, na.rm = TRUE), 2)
)
```

## Student-Level Analysis

The student-level master carries classroom and budget context onto each
student-year row. Use those fields for descriptive summaries and modeling
inputs, not as automatic causal evidence.

```{r student-analysis-preview}
sl <- master$student_level

head(sl[c(
  "school_year",
  "classroom_code",
  "delivery_type",
  "poverty_dum",
  "days_absent_total",
  "gold_literacy_gain_raw",
  "grand_total"
)])
```

```{r student-analysis-summary}
aggregate(
  gold_literacy_gain_raw ~ school_year + delivery_type,
  data = sl,
  FUN = function(x) round(mean(x, na.rm = TRUE), 2)
)
```

## Private Real-Data Linkage

In a private project, build panels as shown in A2, then link and validate them.
Real-data chunks stay non-executed in public documentation.

```{r private-linkage, eval = FALSE}
output_dir <- Sys.getenv("ALPREKDB_OUTPUT_DIR")

master <- linkage_create_master(
  budget = budget_processed$panel,
  classroom = classroom_processed$panel,
  student = student_transform(student_processed$panel)
)

validation <- linkage_validate(master)
coverage <- master$meta$coverage$by_year

validation
coverage
```

Row-level exports should be written only in a private workflow with explicit
local opt-in. For master CSV export, pass a base filename, not a bare directory:
the exporter writes `_classroom.csv` and `_student.csv` suffixes.

```{r private-export, eval = FALSE}
if (identical(Sys.getenv("ALPREKDB_WRITE_OUTPUTS"), "1")) {
  linkage_export_rds(
    master,
    file.path(output_dir, "linkage_master.rds")
  )

  linkage_export_csv(
    master,
    file.path(output_dir, "linkage_master.csv")
  )
}
```

## Next Step

A4 shows how to persist panels and master objects in DuckDB and query them with
SQL.
