---
title: "A4 - DuckDB and SQL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A4 - DuckDB and SQL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_duckdb <- requireNamespace("duckdb", quietly = TRUE) &&
  requireNamespace("DBI", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ALprekDB can persist panel objects and linked master datasets in DuckDB. This
is useful when private workflows need SQL queries, fast reloads, or a stable
local analysis database.

Public examples in this vignette use synthetic data and write only to a
temporary database under `tempdir()`. Real DuckDB files built from ADECE data
are row-level private outputs and should stay outside GitHub, pkgdown, and
package builds.

The database features require the optional `duckdb` and `DBI` packages.

```{r load}
library(ALprekDB)
```

```{r duckdb-missing-note, eval = !has_duckdb, echo = FALSE, results = "asis"}
cat("*DuckDB examples are skipped because optional packages `duckdb` and `DBI` are not installed.*")
```

## Build Synthetic Linked Data

The synthetic setup mirrors A3: budget covers two years while classroom and
student panels cover three years. The third year demonstrates missing budget
coverage through the database round trip.

```{r synthetic-panels, eval = has_duckdb}
n_classrooms <- 12
n_students <- 60
seed <- 42

budget_panel <- alprek_synthetic_budget(
  n_classrooms = n_classrooms,
  n_years = 2,
  seed = seed
)

classroom_panel <- alprek_synthetic_classroom(
  n_classrooms = n_classrooms,
  n_years = 3,
  seed = seed
)

student_panel <- alprek_synthetic_student(
  n_students = n_students,
  n_classrooms = n_classrooms,
  n_years = 3,
  seed = seed
)

student_panel <- student_transform(student_panel)

master <- linkage_create_master(
  budget = budget_panel,
  classroom = classroom_panel,
  student = student_panel
)
```

## Create a Temporary Database

Use a temporary path for examples. Private projects should use a controlled
output directory instead.

```{r create-db, eval = has_duckdb, message = FALSE}
db_path <- tempfile(fileext = ".duckdb")
conn <- db_init(db_path)
```

## Write Panels and Master Tables

`db_write_panel()` chooses a table name from the panel object's S3 class.
`db_write_master()` writes the classroom-level and student-level master tables.

```{r write-db, eval = has_duckdb}
db_write_panel(conn, budget_panel)
db_write_panel(conn, classroom_panel)
db_write_panel(conn, student_panel)
db_write_master(conn, master)

db_list_tables(conn)
```

```{r db-table-assertions, include = FALSE, eval = has_duckdb}
stopifnot(all(c(
  "budget_panel",
  "classroom_panel",
  "student_panel",
  "master_classroom",
  "master_student"
) %in% db_list_tables(conn)))
```

## Inspect Schema and R Types

ALprekDB stores R type metadata in an internal registry. Factor columns are
stored as character values in DuckDB and reconstructed as factors when read
back into R.

```{r inspect-schema, eval = has_duckdb}
head(db_table_info(conn, "classroom_panel"), 10)
```

```{r inspect-master-schema, eval = has_duckdb}
head(db_table_info(conn, "master_student"), 10)
```

## Query with SQL

Use `db_query()` for SQL analysis. It returns a tibble.

```{r sql-budget-year, eval = has_duckdb}
db_query(conn, "
  SELECT
    school_year,
    COUNT(*) AS n_classrooms,
    ROUND(AVG(grand_total), 0) AS mean_budget,
    ROUND(AVG(per_child_budget), 0) AS mean_per_child_budget
  FROM master_classroom
  GROUP BY school_year
  ORDER BY school_year
")
```

The missing budget year remains present, with budget means returned as `NA`.

```{r sql-missing-budget, eval = has_duckdb}
db_query(conn, "
  SELECT
    school_year,
    COUNT(*) AS n_classrooms,
    SUM(CASE WHEN grand_total IS NULL THEN 1 ELSE 0 END) AS n_without_budget
  FROM master_classroom
  GROUP BY school_year
  ORDER BY school_year
")
```

Student-level tables can be queried directly.

```{r sql-student-summary, eval = has_duckdb}
db_query(conn, "
  SELECT
    school_year,
    COUNT(*) AS n_students,
    ROUND(AVG(poverty_dum) * 100, 1) AS pct_poverty,
    ROUND(AVG(days_absent_total), 1) AS mean_absences
  FROM student_panel
  GROUP BY school_year
  ORDER BY school_year
")
```

The linked student master can support grouped descriptive summaries.

```{r sql-linked-students, eval = has_duckdb}
db_query(conn, "
  SELECT
    school_year,
    delivery_type,
    COUNT(*) AS n_students,
    ROUND(AVG(gold_literacy_gain_raw), 1) AS mean_literacy_gain
  FROM master_student
  GROUP BY school_year, delivery_type
  ORDER BY school_year, delivery_type
  LIMIT 10
")
```

## Read Objects Back into R

The read helpers reconstruct ALprekDB S3 objects.

```{r read-back, eval = has_duckdb}
budget_from_db <- db_read_panel(conn, "budget")
classroom_from_db <- db_read_panel(conn, "classroom")
student_from_db <- db_read_panel(conn, "student")
master_from_db <- db_read_master(conn)

data.frame(
  object = c(
    "budget_panel",
    "classroom_panel",
    "student_panel",
    "master_classroom",
    "master_student"
  ),
  original_rows = c(
    nrow(budget_panel$data),
    nrow(classroom_panel$data),
    nrow(student_panel$data),
    nrow(master$classroom_level),
    nrow(master$student_level)
  ),
  read_rows = c(
    nrow(budget_from_db$data),
    nrow(classroom_from_db$data),
    nrow(student_from_db$data),
    nrow(master_from_db$classroom_level),
    nrow(master_from_db$student_level)
  )
)
```

Type reconstruction is part of the round trip.

```{r type-roundtrip, eval = has_duckdb}
data.frame(
  check = c(
    "budget year is integer",
    "classroom delivery_type is factor",
    "student dob is Date",
    "master student dob is Date"
  ),
  passed = c(
    is.integer(budget_from_db$data$year),
    is.factor(classroom_from_db$data$delivery_type),
    inherits(student_from_db$data$dob, "Date"),
    inherits(master_from_db$student_level$dob, "Date")
  ),
  stringsAsFactors = FALSE
)
```

```{r type-assertions, include = FALSE, eval = has_duckdb}
stopifnot(inherits(budget_from_db, "alprek_budget_panel"))
stopifnot(inherits(classroom_from_db, "alprek_classroom_panel"))
stopifnot(inherits(student_from_db, "alprek_student_panel"))
stopifnot(inherits(master_from_db, "alprek_linkage_master"))
stopifnot(is.integer(budget_from_db$data$year))
stopifnot(is.factor(classroom_from_db$data$delivery_type))
stopifnot(inherits(student_from_db$data$dob, "Date"))
stopifnot(inherits(master_from_db$student_level$dob, "Date"))
```

## Reconstruct Coverage Diagnostics

`db_read_master()` reconstructs validation-compatible coverage diagnostics from
the stored master tables.

```{r coverage-roundtrip, eval = has_duckdb}
master_from_db$meta$coverage$by_year[c(
  "school_year",
  "has_budget",
  "has_classroom",
  "has_student",
  "n_budget_rows",
  "n_classroom_rows",
  "n_student_rows",
  "budget_status"
)]
```

```{r coverage-validation, eval = has_duckdb}
db_validation <- linkage_validate(master_from_db)
db_validation
```

```{r coverage-assertions, include = FALSE, eval = has_duckdb}
stopifnot("2023-2024" %in% master_from_db$meta$coverage$missing_budget_years)
stopifnot(any(master_from_db$meta$coverage$by_year$budget_status == "missing_budget"))
stopifnot(db_validation$passed)
```

There is one important boundary: `db_read_master()` rebuilds diagnostics from
the stored `master_classroom` and `master_student` tables. It cannot recover
budget-only right-side rows that were not present in the master tables. If a
future workflow needs to audit those rows after reload, keep the source budget
panel or persisted diagnostics alongside the master tables.

## Year Filters

Read helpers can filter to one or more school years.

```{r read-year-filter, eval = has_duckdb}
classroom_2023 <- db_read_panel(conn, "classroom", years = "2023-2024")
master_2023 <- db_read_master(conn, years = "2023-2024")

data.frame(
  object = c("classroom_2023", "master_classroom_2023", "master_student_2023"),
  rows = c(
    nrow(classroom_2023$data),
    nrow(master_2023$classroom_level),
    nrow(master_2023$student_level)
  )
)
```

```{r read-year-filter-validation, eval = has_duckdb}
linkage_validate(master_2023)
```

## Read-Only Reopen

Once a database exists, it can be opened in read-only mode for analysis.

```{r read-only-reopen, eval = has_duckdb, message = FALSE}
db_close(conn)

conn <- db_init(db_path, read_only = TRUE)

db_query(conn, "
  SELECT COUNT(*) AS n_students
  FROM master_student
")
```

## Private Real-Data DuckDB Files

Real-data DuckDB files are row-level private outputs. Write them only in a
private project and only when row-level output is intentional.

```{r private-duckdb, eval = FALSE}
output_dir <- Sys.getenv("ALPREKDB_OUTPUT_DIR")
private_db_path <- file.path(output_dir, "alprekdb.duckdb")

if (identical(Sys.getenv("ALPREKDB_RUN_REALDATA"), "1") &&
    identical(Sys.getenv("ALPREKDB_WRITE_OUTPUTS"), "1")) {
  conn <- db_init(private_db_path)
  on.exit(db_close(conn), add = TRUE)

  db_write_panel(conn, budget_processed$panel, overwrite = TRUE)
  db_write_panel(conn, classroom_processed$panel, overwrite = TRUE)
  db_write_panel(conn, student_transform(student_processed$panel), overwrite = TRUE)
  db_write_master(conn, master, overwrite = TRUE)
}
```

Do not commit `.duckdb` files, row-level exports, `_targets/` caches, or
private output folders. Publish aggregate diagnostics and non-disclosive counts
instead.

## Cleanup

```{r cleanup, include = FALSE, eval = has_duckdb}
db_close(conn)
unlink(db_path)
```

## Next Step

A5 shows how to run the package's companion `targets` workflow for repeatable
synthetic and private real-data processing.
