---
title: "A7: Geocoding Quality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A7: Geocoding Quality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## 1. Overview

The **geocode** module (added in v0.8.0) reconciles two coordinate
sources for each Alabama First Class Pre-K site:

- **Melissa.com** — commercial geocoding vendor (delivered as a 29-column
  Excel file each release cycle). Melissa returns site latitude /
  longitude plus a `RESULTCODE` describing the precision tier of every
  lookup.
- **ADECE** — administrative coordinates that ADECE staff maintain in
  their internal classroom data (carried through the
  `alprek_classroom_panel`'s `latitude` / `longitude` columns).

Both sources have known weaknesses. Melissa can fall back to a ZIP
centroid when an address is unparseable. ADECE coordinates can drift if
a site moves or if the original entry was a Google-Maps-on-iPhone tap
rather than a careful rooftop pin. The geocode module's job is to
**pick one authoritative coordinate per site-year**, **explain why**,
and **flag the rows that need analyst follow-up**.

This vignette walks through that pipeline end-to-end on **synthetic
data** (`alprek_synthetic_geocode()`), so every chunk runs without an
ADECE/Melissa file. Real-data calls look identical — just point
`geocode_read()` at the actual Melissa `.xlsx`.

```{r load}
library(ALprekDB)
```

The module follows the same shape as every other ALprekDB module: read
-> detect -> clean -> validate -> reconcile -> transform -> panel ->
export -> linkage -> DuckDB persistence.

## 2. Scope (v0.8.0)

**In-scope** (this module):

- Single Melissa snapshot (the `2026-03-04` v1 delivery)
- 7 codebooks: `geocode_column_map_melissa_v1`,
  `melissa_resultcode_codes`, `melissa_statuscode_codes`,
  `melissa_errorcode_codes`, `geocode_al_fips_counties`,
  `geocode_source_manifest`, `geocode_edge_cases`
- 15-cell decision matrix for picking an authoritative coordinate
- Per-tier distance thresholds (50 m / 250 m / 500 m / Inf) from
  observed Melissa precision medians
- 10 authoritative reconcile columns (incl. `coord_model_status`)
- Multi-format export (CSV / Parquet / Excel / RDS / Stata)
- Follow-up queue CSV exporter
- Linkage to `classroom_panel` and `applications_master`
- DuckDB persistence (4 new tables)

**Out-of-scope** (intentional; mirrors the v0.7.0 stance retraction in
NEWS):

- ACS area-weighted aggregation -> future `ALprekACS`
- OSRM isochrone / drive-time -> future package
- Bayesian small-area estimation -> future `ALprekSAE`
- Live vendor API calls — permanent out-of-scope
- Multi-source consensus (ArcGIS / Google fallback) — future release

The acceptability of any Melissa `RESULTCODE` for SAE modeling is
governed by the **codebook**
(`inst/extdata/codebooks/melissa_resultcode_codes.csv`,
`acceptable_for_master` column), not by Melissa's vendor documentation.
The decision matrix and reconciler read that codebook through
`alprek_geocode_resultcode_meaning()` — there is **no hardcoded vendor
semantics in R/**. To change a tier's acceptability, edit the CSV and
the package picks it up.

## 3. Read the Melissa file

`geocode_read()` reads the Melissa-returned xlsx (default sheet:
`Sheet1`), stamps `raw_row_index` and a row-stable `lineage_id`, and
captures full provenance (file SHA-256, git SHA, cycle year, receipt
date) in `$meta`.

```{r read-real, eval = FALSE}
raw <- geocode_read(
  path         = file.path("ORIGINAL-DATA",
                            "2026-03-04_Pre-K Geocoding Melissa",
                            "2026-03-04_geocoding_master_Final.xlsx"),
  cycle_year   = "2026-2027",
  receipt_date = "2026-03-04"
)
raw
#> <alprek_geocode_raw>
#>   path:          .../2026-03-04_geocoding_master_Final.xlsx
#>   cycle_year:    2026-2027
#>   receipt_date:  2026-03-04
#>   n_rows: 3396  n_cols: 29
```

For the rest of this vignette we use synthetic data of identical
shape:

```{r read-synthetic}
g_synth <- alprek_synthetic_geocode(
  n_sites = 30L,
  n_years = 3L,
  seed    = 42L
)
dim(g_synth)
names(g_synth)[1:12]
```

`alprek_synthetic_geocode()` returns a flat 29-column tibble that
matches the Melissa v1 contract exactly. To pass it through the rest
of the pipeline we wrap it in an `alprek_geocode_raw` shell (this
mirrors `geocode_read()`'s output and is the same approach the test
suite uses):

```{r synth-shell}
fake_geocode_raw <- function(df,
                              cycle_year   = "2026-2027",
                              receipt_date = "2026-03-04") {
  # Mirror the dtype contract that readxl produces on a real v1 file.
  df$LAT        <- as.character(df$LAT)
  df$LNG        <- as.character(df$LNG)
  df$ERRORCODE  <- as.logical(df$ERRORCODE)
  df$site_zip   <- as.numeric(df$site_zip)
  df$COUNTYNAME <- toupper(df$COUNTYNAME)
  df$raw_row_index <- seq_len(nrow(df))
  df$lineage_id    <- sprintf("synth_%06d", df$raw_row_index)
  meta <- list(
    path          = "/tmp/fake-melissa.xlsx",
    sheet         = "Sheet1",
    cycle_year    = cycle_year,
    receipt_date  = receipt_date,
    source        = "melissa",
    file_sha256   = paste(rep("a", 64L), collapse = ""),
    file_basename = "synthetic-melissa.xlsx",
    git_sha       = "synthetic",
    n_rows        = nrow(df),
    n_cols        = ncol(df) - 2L,
    col_names     = setdiff(names(df), c("raw_row_index", "lineage_id")),
    read_at       = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"),
    lineage_id    = df$lineage_id,
    raw_row_index = df$raw_row_index
  )
  structure(list(data = tibble::as_tibble(df), meta = meta),
            class = "alprek_geocode_raw")
}

raw <- fake_geocode_raw(g_synth)
raw
```

## 4. Detect, clean, validate

### Detect

`geocode_detect_format()` examines marker columns (`RESULTCODE`,
`STATUSCODE`, the 29-column contract) to classify the input as
`"v1"` (the 2026-03-04 contract), `"unknown_with_overlap"`, or
`"unknown"`. This is the seam future Melissa deliveries plug into.

```{r detect}
fmt <- geocode_detect_format(raw)
fmt$format
fmt$confidence
```

### Clean

`geocode_clean()` runs an 11-step pipeline: applies the column map,
coerces `LAT` / `LNG` from character to numeric (the **key**
transformation), coerces `ERRORCODE` to character, preserves
ZIP-family fields, validates `school_year`, title-cases
`COUNTYNAME`, trims whitespace, and attaches a `data_source_map`
attribute (ADECE vs. Melissa-2026-03) on every column. `lineage_id`
is preserved unchanged.

```{r clean}
clean <- geocode_clean(raw)
clean
```

The `$cleaning_log` records every rule applied, with `INFO` / `WARN` /
`ERROR` severity:

```{r cleaning-log}
clean$cleaning_log[, c("rule", "n_affected", "severity")]
```

### Validate

`geocode_validate()` runs 15 checks against the codebook contracts.
Real-data v1 (3,396 rows) typically returns `passed = TRUE` with a
handful of WARN-level rows for known-quirky records.

```{r validate}
v <- geocode_validate(clean)
v
```

## 5. The decision matrix

The reconciler picks an authoritative `(lat, lng)` pair per row using
a **15-cell decision matrix**. Each cell is keyed on three inputs:

1. Is ADECE present? (`latitude` / `longitude` non-NA)
2. Is Melissa present? (`LAT` / `LNG` non-NA after numeric coercion)
3. Melissa `RESULTCODE` (precision tier; see codebook below)

The four Melissa codes observed in v1 are:

| `RESULTCODE` | Melissa label                   | Precision tier in `lat_precision` | Per-tier threshold |
|---|---|---|---|
| `GS01` | Street level (ZIP+4 / Plus4)       | `zip4`     | 50 m   |
| `GS05` | Rooftop within property boundaries | `rooftop`  | 250 m  |
| `GS06` | Interpolated rooftop               | `parcel`   | 500 m  |
| `GS03` | Community / ZIP centroid           | `zip5`     | Inf (always flag) |

These labels are vendor-accurate per Melissa's *Result Code Details*
documentation. The first cycle of this module (pre-Step 2.1) used
inverted labels; the v0.8.0 codebook is the corrected reference.

### The 15 cells (Step 4.2 LOCKED)

| Cell | ADECE | Melissa | `RESULTCODE` | Distance band       | Outcome |
|---|---|---|---|---|---|
| D1  | Y | Y | `GS01` | <=50 m            | melissa, zip4, no follow-up |
| D2  | Y | Y | `GS01` | 50 m - 10 km      | melissa, zip4, flag (`disagreement_above_threshold`) |
| D3  | Y | Y | `GS01` | >=10 km           | `disputed_melissa`, flag (`disagreement_gross`) |
| D4  | Y | Y | `GS05` | <=250 m           | melissa, rooftop, no follow-up |
| D5  | Y | Y | `GS05` | 250 m - 10 km     | melissa, rooftop, flag |
| D6  | Y | Y | `GS05` | >=10 km           | `disputed_melissa`, flag (gross) |
| D7  | Y | Y | `GS06` | <=500 m           | melissa, parcel, no follow-up |
| D8  | Y | Y | `GS06` | 500 m - 10 km     | melissa, parcel, flag |
| D9  | Y | Y | `GS06` | >=10 km           | `disputed_melissa`, flag (gross) |
| D10 | Y | Y | `GS03` | any               | `disputed_melissa`, flag (`resultcode_gs03_always_flag`) |
| D11 | Y | N | --     | --                | adece, flag (`melissa_unexpectedly_missing`) |
| D12 | N | Y | `GS01`/`GS05` | --         | melissa, no flag |
| D13 | N | Y | `GS06` | --                | melissa, flag (`melissa_only_interpolated`) |
| D14 | N | Y | `GS03` | --                | melissa, flag (`melissa_only_gs03`) |
| D15 | N | N | --     | --                | none, flag (`both_missing`) |

`GS03` (ZIP centroid) is **always** flagged because a ZIP centroid is
not a site coordinate even when it happens to land near the ADECE
coordinate. Empirically, `GS03` rows in the v1 delivery have a median
ADECE-Melissa disagreement of ~4 km; treating them as model-ready
would inject ZIP-grain noise into a site-grain estimator.

You can swap the per-tier scheme for a flat threshold:

```{r config-options, eval = FALSE}
cfg_flat <- geocode_config(
  path                     = "...",
  cycle_year               = "2026-2027",
  delivery_date            = "2026-03-04",
  distance_threshold_rules = "flat_250m"
)
```

`by_resultcode` is the default (and what we use for the rest of this
vignette).

## 6. Reconcile

`geocode_reconcile()` walks every row through the matrix, computes the
ADECE-Melissa haversine distance when both are present, and emits 10
new authoritative columns.

```{r reconcile}
rec <- geocode_reconcile(clean)
rec
```

The 10 authoritative columns added to `$data`:

| Column | Type | Meaning |
|---|---|---|
| `lat_final` / `lng_final`    | numeric         | The chosen coordinate (NA where matrix cell == D15) |
| `lat_source`                 | factor          | `{melissa, adece, disputed_melissa, none}` |
| `lat_precision`              | ordered factor  | `{none, unknown, centroid, zip5, zip4, area, parcel, rooftop}` |
| `distance_adece_melissa_m`   | numeric         | Haversine distance, NA when only one source |
| `coord_agreement_band`       | factor          | `{exact, tight, loose, drift, gross, one_source_only, none}` |
| `needs_followup_geocoding`   | logical         | The analyst-queue flag |
| `followup_reason`            | factor          | Controlled vocabulary (see section 13) |
| `coord_model_status`         | ordered factor  | `{missing, not_model_ready, provisional_followup, model_ready}` |
| `geocode_provenance`         | character       | Compact audit string per row |

The `coord_model_status` column is the **SAE-readiness gate** finalized
for v0.8.0. Only `model_ready` rows should feed an
unqualified SAE estimator; `provisional_followup` and
`not_model_ready` rows remain visible in every export so downstream
consumers can choose to downweight or exclude them — they are
**never silently dropped**.

```{r inspect-reconciled}
rec$data[, c("row_id", "lat_source", "lat_precision",
              "distance_adece_melissa_m", "coord_agreement_band",
              "needs_followup_geocoding", "followup_reason",
              "coord_model_status")] |>
  head(8)
```

The `$reconciliation_log` records the matrix cell, threshold used, and
a textual note for every input row — the per-row audit chain analysts
need when reviewing a decision:

```{r reconcile-log}
rec$reconciliation_log[, c("row_id", "matrix_cell", "lat_source",
                            "distance_adece_melissa_m",
                            "threshold_name", "followup_reason")] |>
  head(5)
```

`$summary` gives a per-cell count for the run:

```{r reconcile-summary}
rec$summary
```

## 7. Transform: 5 derived variables

`geocode_transform()` adds **data-layer** derived variables (no
geocoding API calls, no ACS, no Bayesian — those are out-of-scope).
The result is an `alprek_geocode_master`.

```{r transform}
mst <- geocode_transform(rec)
mst
```

Derived variables:

| Variable | Type | Notes |
|---|---|---|
| `precision_tier`     | ordered factor | Aliased from `lat_precision`, levels reversed so `sort()` puts rooftop rows first |
| `in_alabama`         | logical NA-able | `lat_final` in [30, 36] AND `lng_final` in [-89, -84] |
| `county_check_match` | logical NA-able | Melissa `COUNTYNAME` vs. an ADECE county sidecar when supplied; otherwise `NA` |
| `coord_age_years`    | integer NA-able | `cycle_year_first - school_year_first` |
| `geocode_run_id`     | character       | Panel-stable identifier, e.g., `"melissa_v1_2026-03"` |

```{r derived-log}
mst$transform_log[, c("rule", "n_affected", "severity")]
```

The `$data` slot preserves `lineage_id` (Step 3.1) and
`coord_model_status` (Step 4.3) untouched. Exporters in section 9
respect that contract: rows with `coord_model_status != "model_ready"`
are **visible** in every output, not silently filtered.

## 8. Multi-cycle panel

`geocode_bind_years()` stacks multiple `alprek_geocode_master`
snapshots into an `alprek_geocode_panel`. **Important**: a single
Melissa delivery is already a 5-year long panel (school_year ∈
`{2021-2022, ..., 2025-2026_new}`); the within-delivery long shape is
materialized upstream. `geocode_bind_years()` is for binding
**multiple Melissa runs across release cycles** (e.g., a future v0.9.0
delivery on top of the current v0.8.0 delivery). For v0.8.0 with a
single run, the call collapses to a degenerate identity that returns a
1-run panel whose `$data` equals the input's `$data`.

```{r bind-years}
panel <- geocode_bind_years(mst)
panel
```

The binding log records one row per run with file SHA-256 and snapshot
date:

```{r binding-log}
panel$binding_log
```

The unique key in the bound panel is `(row_id, geocode_run_id)` — the
same `row_id` is allowed across runs (a renewal site re-geocoded each
cycle).

## 9. Export (5 formats + the follow-up queue)

Standard data exports preserve every row (including
`not_model_ready` and `provisional_followup` rows):

```{r export, eval = FALSE}
geocode_export_csv(mst)
#> -> "output/geocode/geocode_melissa_v1_2026-03.csv"
geocode_export_parquet(mst)
geocode_export_excel(mst,   include_summary = TRUE)
geocode_export_rds(mst)
geocode_export_stata(mst)
```

The Excel exporter optionally writes a `Summary` sheet with the
`coord_model_status` / `lat_source` distributions and the count of
rows flagged `needs_followup_geocoding`.

## 10. Linkage with classroom + applications

The geocode module hooks into the existing linkage layer at two
points.

### `linkage_geocode_classroom()`

Joins an `alprek_geocode_panel` onto an `alprek_classroom_panel` so
every classroom-year row inherits the per-site authoritative
coordinates. Geocoding happens at the **site** grain; the classroom
panel is at classroom-year grain, so multiple classrooms at the same
site share one geocode row. The function preserves classroom row
order and never inflates.

```{r linkage-classroom, eval = FALSE}
panel_c <- classroom_bind_years(c2122, c2223, c2324, c2425)
lk      <- linkage_geocode_classroom(panel, panel_c)
lk
#> <alprek_geocode_linkage_classroom>
#>   Classroom rows:       2410
#>   Geocode rows (in):    3396
#>   Matched (classroom):  2287 (94.9%)
#>   Unmatched classroom:  123
#>   Unmatched geocode:    1232
```

The 12 attached columns are prefixed `geocode_*` (e.g.,
`geocode_lat_final`, `geocode_coord_model_status`,
`geocode_lineage_id`, `geocode_run_id`) so they cannot collide with
the classroom panel's ADECE `latitude` / `longitude` columns —
those are kept verbatim as an escape hatch for inspection.

### `linkage_geocode_applications()`

Joins the geocode panel onto an `alprek_applications_master`. For
renewals + bucket-C new applications, joins on
`(matched_site_code, school_year)`. For bucket-D "truly new"
applications (no `site_code` yet), joins on `row_id` — the Melissa
file's `2025-2026_new_NNNN` row_ids correspond directly to
bucket-D applications.

```{r linkage-apps, eval = FALSE}
lk_apps <- linkage_geocode_applications(panel, app_master)
lk_apps$diagnostics
```

## 11. `linkage_create_master()` — the Goal #2 deliverable

The master extension (v0.8.0) gives `linkage_create_master()` two new
optional arguments: `geocode = ` and `applications = `. The signature
is **backward compatible** — a 3-arg call produces v0.7.0 output.

```{r create-master, eval = FALSE}
# v0.7.0 (3-arg) — unchanged
master <- linkage_create_master(budget_panel, classroom_panel,
                                  student_panel)

# v0.8.0 with reconciled geocode
master <- linkage_create_master(
  budget_panel, classroom_panel, student_panel,
  geocode = panel
)

# Full v0.8.0 with geocode + applications
master <- linkage_create_master(
  budget_panel, classroom_panel, student_panel,
  geocode      = panel,
  applications = app_master
)
master$classroom_level |>
  dplyr::select(school_year, classroom_code, site_code,
                  geocode_lat_final, geocode_lat_source,
                  geocode_coord_model_status,
                  geocode_needs_followup_geocoding) |>
  head(5)
```

`master$classroom_level` gains 12 prefixed `geocode_*` columns: the
10 authoritative/model-readiness columns plus `geocode_run_id` and
`geocode_lineage_id`. The ADECE `latitude` /
`longitude` columns remain untouched. `master$diagnostics` includes
`geocode_coverage` (rolled-up percentage of model-ready coords) and
`geocode_linkage` (the underlying diagnostic tibble).

## 11.5 Analysis-time merge patterns

By design, the module-level production panels at `output/budget/`,
`output/classroom/`, `output/student/`, and `output/applications/`
do **not** carry the geocode columns. ALprekDB keeps cleaning and
validation strictly module-local: each panel is the canonical output
of one cleaning pipeline. Joining geocode to a panel is the analyst's
choice, exercised at analysis time using the package's `linkage_*()`
functions. The benefit is straight separation of concerns: a budget
cleaning bug never has to wait on a geocoding rebuild, and analysts
who need a different cycle of Melissa data can swap it in freely.

Four common patterns follow. Each is 2-4 lines of code.

**Pattern A — Classroom + geocode.** Direct join on
`(site_code, school_year)`. Adds 12 prefixed `geocode_*` columns
to the classroom panel.

```{r merge-classroom, eval = FALSE}
classroom <- readRDS("output/classroom/classroom_panel_2021-2026.rds")
# panel_geo is the alprek_geocode_panel built earlier in this vignette
linked <- linkage_geocode_classroom(panel_geo, classroom)
linked$data  # 7,409 rows + 12 geocode_* cols + ADECE latitude/longitude preserved
```

**Pattern B — Applications + geocode.** Two-phase join:
`matched_site_code` for renewals (buckets A/B/C) and `row_id` for
bucket-D "truly new" applications whose `_new_NNNN` row_ids correspond
directly to Melissa rows.

```{r merge-applications, eval = FALSE}
apps   <- readRDS("output/applications/applications_master_2026-2027.rds")
linked <- linkage_geocode_applications(panel_geo, apps)
linked$data  # 1,617 rows + 12 geocode_* cols, with bucket-D rows joined on row_id
```

**Pattern C — Budget / student panels.** Budget has only
`classroom_code` (no `site_code`), so `linkage_create_master()`
handles the indirect resolution. Student is at student-year grain;
the geocode value is propagated via the classroom join inside
the master integrator.

```{r merge-budget-student, eval = FALSE}
budget    <- readRDS("output/budget/budget_panel_2021-2025.rds")
student   <- readRDS("output/student/student_panel_2021-2026.rds")
m <- linkage_create_master(budget, classroom, student, geocode = panel_geo)
m$classroom_level  # budget cols + classroom cols + student aggregates + 12 geocode_* cols
m$student_level    # 116,689 student-year rows with classroom-derived geocode_* cols
```

**Pattern D — Full integrated master.** This is the same call
that produced
`output/master/master_classroom_level_2026-2027_v0.8.0.csv`
in the v0.8.0 release packet; rerun it locally to inspect alongside
the other panels.

```{r merge-master, eval = FALSE}
apps <- readRDS("output/applications/applications_master_2026-2027.rds")
m <- linkage_create_master(budget, classroom, student,
                           geocode      = panel_geo,
                           applications = apps)
m$classroom_level  # 7,409 rows × 246 cols: budget + classroom + student-agg + 12 geocode_* + 19 app_*
```

The geocode panel itself can be loaded from the v0.8.0 release
artifact or rebuilt from a Melissa xlsx; both flows are equivalent.

```{r geocode-reload, eval = FALSE}
# Option 1: Reload from release artifact
rec <- readRDS("output/geocode/geocode_reconciled_2026-2027.rds")
master_geo <- geocode_transform(rec, config = geocode_config(
  path = rec$meta$path, cycle_year = rec$meta$cycle_year,
  delivery_date = rec$meta$receipt_date, verbose = FALSE))
panel_geo <- geocode_bind_years(master_geo)

# Option 2: Rebuild from xlsx (Sections 3-8 of this vignette)
# raw <- geocode_read(...); clean <- geocode_clean(raw); ...
```

## 12. DuckDB persistence

For long-term storage and SQL access, the geocode module writes 4
tables into the shared DuckDB file alongside budget / classroom /
student / applications:

- `geocode_clean`       — per `geocode_run_id`, cleaned standardized cols
- `geocode_reconciled`  — per run, full reconciled rows + 10 auth cols + `lineage_id`
- `geocode_panel`       — cross-run, long-format multi-snapshot view
- `geocode_lineage`     — one row per write, capturing run-level meta (file SHA-256, n_rows, n_followup, threshold rules, timestamps)

```{r duckdb, eval = FALSE}
conn <- db_init("output/alprekdb.duckdb")

# Per-run write
db_write_geocode_clean(conn, clean)
db_write_geocode_reconciled(conn, rec)
db_write_geocode_panel(conn, panel, overwrite = TRUE)

# Round-trip
rec_back   <- db_read_geocode_reconciled(conn)
panel_back <- db_read_geocode_panel(conn)
lineage_df <- db_read_geocode_lineage(conn)

DBI::dbDisconnect(conn, shutdown = TRUE)
```

Ordered-factor levels for `lat_precision`, `coord_model_status`, and
`precision_tier` round-trip through the shared column-type registry.

## 13. The follow-up queue (Goal #3)

`geocode_export_followup_queue()` is the user-facing surface for the
analyst-driven re-geocoding workflow. It surfaces every row flagged
`needs_followup_geocoding = TRUE`, annotates each with a controlled
`suggested_action`, and writes
`output/geocode/sites_needing_geocoding_<cycle>.csv`.

```{r followup-queue}
fq <- geocode_followup_queue(rec)
nrow(fq)
fq[, c("row_id", "school_year", "lat_source",
        "distance_adece_melissa_m", "melissa_result_code",
        "followup_reason", "suggested_action")] |>
  head(5)
```

The disk-bound exporter behaves the same and adds an internal-use
header on the CSV:

```{r followup-export, eval = FALSE}
fq <- geocode_export_followup_queue(
  rec,
  cycle_year       = "2026-2027",
  internal_use     = TRUE,   # default — adds "DO NOT REDISTRIBUTE" header
  include_disputed = TRUE
)
attr(fq, "output_path")
#> "output/geocode/sites_needing_geocoding_2026-2027.csv"
attr(fq, "privacy_level")
#> "internal_address_followup"
```

### Interpreting `followup_reason` -> `suggested_action`

The controlled vocabulary mapping is **deterministic** so a downstream
caller can group / filter without re-reading the decision matrix:

| `followup_reason`                       | `suggested_action`                              | What to do |
|---|---|---|
| `disagreement_above_threshold`          | `manual_source_adjudication`                    | Both sources present, distance > tier threshold but < 10 km. Inspect both addresses; usually one is the program site, the other a billing or mailing address. |
| `disagreement_gross`                    | `verify_adece_address_and_request_recheck`      | Distance >= 10 km. Usually an ADECE address-field error. Check ADECE first, then re-request the Melissa run if needed. |
| `resultcode_not_acceptable_for_master`  | `request_higher_precision_geocode`              | Vendor RESULTCODE is in the codebook but `acceptable_for_master = FALSE` (e.g., `GS02`/`GS04` if future Melissa runs return them). |
| `both_missing`                          | `obtain_coord`                                  | Neither ADECE nor Melissa produced a coordinate. Obtain coordinate by any means. |
| `melissa_unexpectedly_missing`          | `request_melissa_geocode`                       | ADECE has a coord, Melissa returned blank. Re-submit to Melissa. |
| `melissa_only_interpolated`             | `request_higher_precision_geocode`              | Melissa returned `GS06` interpolated rooftop and ADECE is absent. Re-geocode at higher precision. |
| `melissa_only_gs03`                     | `request_higher_precision_geocode`              | Melissa returned `GS03` ZIP centroid and ADECE is absent. Re-geocode. |
| `resultcode_gs03_always_flag`           | `manual_review_gs03`                            | `GS03` ZIP centroid with ADECE present. Decide whether ADECE is reliable enough to override the centroid. |
| (other / NA)                            | `manual_review`                                 | Catch-all for unmapped reasons. |

Sort order in the exported queue: `school_year` descending, then
`distance_adece_melissa_m` descending (largest disagreements first;
one-source-only rows sort to the end).

## 14. Caveats

### Privacy: `site_street` is a full address

The follow-up queue contains full site addresses (`site_street`,
`site_city`, `site_state`, `site_zip`) because the analyst needs
them to verify / re-geocode. By default the in-memory tibble carries:

- `attr(fq, "privacy_level")        = "internal_address_followup"`
- `attr(fq, "contains_address_fields") = TRUE`
- `attr(fq, "internal_use")         = TRUE` (when `internal_use = TRUE`)

The on-disk CSV is prefixed with a comment header
`# INTERNAL USE -- DO NOT REDISTRIBUTE`. **Do not commit the
follow-up queue to a public repository or share it outside the data
team.** The `output/` and `ORIGINAL-DATA/` directories are excluded
from the package's public mirror by design.

### GS03 is always flagged

Even when a `GS03` row has a Melissa coordinate that lands near the
ADECE coordinate, the matrix flags it. A ZIP centroid is not a site
coordinate; the apparent agreement is artifactual. If ADECE is
present, you can use ADECE as the provisional final coordinate —
that's what `authoritative_priority = "adece_first"` does — but the
row still needs analyst attention.

### Coord age (`coord_age_years`)

ADECE coordinates can be years old. `coord_age_years =
cycle_year_first - school_year_first` is the rough freshness gauge;
large positive values mean the panel row is being modeled against an
older ADECE coordinate. Sites that move (or that grow into a new
classroom at a new street) can break the assumption that an old
coordinate still describes the program. This is informational only;
no automatic correction is applied.

### Centroid vs. rooftop precision

`GS01` (ZIP+4 street level, 50 m threshold) is the **tightest**
threshold in v0.8.0 — much tighter than `GS05` (rooftop, 250 m) and
`GS06` (interpolated rooftop, 500 m). This counterintuitive ordering
comes from the v0.8.0 empirical medians: the small `GS01` sample
clustered tightly, while `GS06`'s interpolation produces a wider
acceptable disagreement window. If a future delivery shifts those
distributions, edit `inst/extdata/codebooks/melissa_resultcode_codes.csv`
and the per-tier list in `geocode_config()`.

### `STATUSCODE` is a derived field

Melissa publishes `RESULTCODE` in its public documentation; the
`STATUSCODE` column is internal to this delivery and not documented
externally. We treat the codebook's 1:1 mapping
(`GS01<->9`, `GS03<->5`, `GS05<->B`, `GS06<->A`) as
**authoritative for this v1 contract** and re-derive validation
against it. Future deliveries that change the mapping should update
`inst/extdata/codebooks/melissa_statuscode_codes.csv` first.

### Acceptability is codebook-governed

`acceptable_for_master` in `melissa_resultcode_codes.csv` is the
**single source of truth** for whether a Melissa RESULTCODE is
SAE-ready. The reconciler reads it through
`alprek_geocode_resultcode_meaning()`. To re-include `GS02`/`GS04` /
exclude `GS06` for a future modeling cycle, edit the CSV — no R
changes required.

## What this module does NOT do

| Concern | Where it lives |
|---|---|
| Live geocoding API calls       | Permanently out-of-scope     |
| Multi-source consensus (ArcGIS, Google fallback) | Future release |
| ACS area-weighted aggregation  | Future `ALprekACS` package    |
| OSRM isochrone / drive-time    | Future package                |
| Bayesian SAE of economic-need tiers | Future `ALprekSAE` package |

This split is intentional: the data-contract layer (this module) must
be stable, audited, and codebook-driven before spatial / statistical
layers consume `lat_final` as truth.

## See also

- `vignette("a1-getting-started", package = "ALprekDB")` — package overview
- `vignette("a3-linkage-analysis", package = "ALprekDB")` — broader linkage
  patterns this module integrates with
- `vignette("a4-duckdb-sql", package = "ALprekDB")` — DuckDB schema + SQL
- `vignette("a6-applications-intake", package = "ALprekDB")` — sibling module
  this module mirrors in pattern
- `vignette("m2-validation-framework", package = "ALprekDB")` — how
  `geocode_validate()` plugs into the package-wide framework
- `vignette("m4-privacy-provenance", package = "ALprekDB")` — full
  treatment of the privacy guardrails the follow-up queue inherits
