---
title: "Getting Started with ALprekDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ALprekDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ALprekDB is a modular R package for processing, cleaning, validating, and
managing Alabama First Class Pre-K (FCPK) administrative data received from
ADECE (Alabama Department of Early Childhood Education).

The package provides a complete pipeline from raw Excel files to analysis-ready
panel datasets:

```
Excel files  -->  Read  -->  Clean  -->  Validate  -->  Panel  -->  Linkage  -->  Database
(Budget)          S3         S3          S3             S3          S3            DuckDB
(Classroom)       objects    objects     objects        objects     objects
(Student)
```

### Package Modules

| Module | Purpose |
|--------|---------|
| **Budget** | Per-classroom funding (OSR and Other sources, 11 categories) |
| **Classroom** | Classroom characteristics, teacher demographics, geography |
| **Student** | Student demographics, assessments (GOLD, PPVT, eDECA, ASQ), services |
| **Linkage** | Cross-module joins to create master datasets |
| **Transform** | Derived variables (gain scores, chronic absence, risk index) |
| **Database** | DuckDB persistent storage with SQL query support |

## Installation

```{r install, eval = FALSE}
# From GitHub:
# install.packages("remotes")
remotes::install_github("joonho112/ALprekDB")
```

```{r load}
library(ALprekDB)
```

## Classroom Codes

Every record in ADECE data is identified by a **classroom code** with the format
`CCCDNNNNN.NN`:

- `CCC` = County code (001-067)
- `D` = Delivery type code (P=Public, C=Community, H=Head Start, etc.)
- `NNNNN` = Program number
- `NN` = Class number within the program

Use `parse_classroom_codes()` to extract these components:

```{r parse-codes}
codes <- c("001P12345.01", "067H54321.02", "033C98765.01")
parsed <- parse_classroom_codes(codes)
parsed
```

## Codebook System

ALprekDB uses data-driven CSV codebooks stored in `inst/extdata/` for
column mappings and value standardization. This makes the cleaning rules
transparent and easily auditable.

```{r codebooks}
# Budget category mappings
budget_categories <- alprek_category_groups()
head(budget_categories, 5)

# Delivery type code mappings
delivery_types <- alprek_delivery_types()
delivery_types

# Race mapping (classroom module)
race_map <- alprek_race_mapping()
race_map
```

## Quick Start with Synthetic Data

For demonstrations, ALprekDB includes synthetic data generators that produce
realistic datasets without requiring access to confidential ADECE files:

```{r synthetic-quick}
# Generate synthetic panels (all share classroom codes for linkage)
budget <- alprek_synthetic_budget(n_classrooms = 15, n_years = 2, seed = 42)
classroom <- alprek_synthetic_classroom(n_classrooms = 15, n_years = 2, seed = 42)
student <- alprek_synthetic_student(n_students = 60, n_classrooms = 15,
                                     n_years = 2, seed = 42)
```

```{r inspect-budget}
budget
head(budget$data[, c("school_year", "classroom_code", "delivery_type",
                      "grand_total", "share_osr")])
```

```{r inspect-student}
student
```

## Real Data Processing Pipeline

With actual ADECE Excel files, the processing workflow follows these steps:

```{r real-pipeline, eval = FALSE}
# 1. Configure
config <- budget_config(
  school_year = "2023-2024",
  path = "path/to/FCPK_Budget_23-24.xlsx"
)

# 2. Process (read + clean + validate)
result <- budget_process(config)

# 3. Inspect
result$raw       # alprek_budget_raw
result$clean     # alprek_budget_master
result$validation  # alprek_budget_validation

# 4. Multi-year panel
configs <- list(
  budget_config("2021-2022", "Budget_21-22.xlsx"),
  budget_config("2022-2023", "Budget_22-23.xlsx"),
  budget_config("2023-2024", "Budget_23-24.xlsx"),
  budget_config("2024-2025", "Budget_24-25.xlsx")
)
panel_result <- budget_process_years(configs)
panel <- panel_result$panel  # alprek_budget_panel
```

## Exploring the Panel

```{r explore-panel}
# Budget summary by year
budget_summary <- budget_summary_stats(budget)
budget_summary[, c("school_year", "n", "grand_total_mean",
                    "share_osr_mean")]
```

## Validation

Every module includes a validation step that checks data quality:

```{r validation-example, eval = FALSE}
# After cleaning
validation <- budget_validate(cleaned_obj, strict = FALSE)
print(validation)
# Shows: checks passed, warnings, errors
```

## Export

Export panel data to multiple formats:

```{r export-example, eval = FALSE}
budget_export_csv(panel, "output/budget_panel.csv")
budget_export_rds(panel, "output/budget_panel.rds")
budget_export_stata(panel, "output/budget_panel.dta")
```

## Next Steps

- **Panel Construction**: See `vignette("panel-construction")` for multi-year
  processing workflows
- **Linkage & Analysis**: See `vignette("linkage-and-analysis")` for
  cross-module linkage and DuckDB integration
- **Architecture**: See `vignette("reference-architecture")` for S3 class
  hierarchy and data dictionary
