---
title: "M2 - Validation Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{M2 - Validation Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

Validation in ALprekDB is structured evidence about processing quality. It is
not a claim that administrative records are perfect, and it is not a substitute
for substantive analysis review. Its job is narrower and more operational:

- make processing assumptions inspectable;
- apply the same checks across repeated runs;
- distinguish data-quality problems from expected coverage gaps;
- report aggregate diagnostics without exposing confidential rows.

This is the validation side of the public-code, private-data architecture
described in M1.

```{r load}
library(ALprekDB)
```

## Validation Objects

Validation functions return S3 objects with a common structure.

| Field | Meaning |
|-------|---------|
| `passed` | Overall result under default or strict interpretation. |
| `n_errors` | Number of `ERROR` checks. |
| `n_warnings` | Number of `WARN` checks. |
| `n_info` | Number of `INFO` checks. |
| `checks` | A tibble with one row per check. |
| `issues` | Non-linkage validators also include specific issue rows when available. |

The `checks` tibble uses a stable set of columns:

```text
check_name
check_description
status
n_issues
details
```

The number of checks depends on the validation object. At the current source
state, the core counts are:

| Validator | Current check count |
|-----------|--------------------:|
| `budget_validate()` | 7 |
| `classroom_validate()` | 10 |
| `student_validate()` | 12 |
| `linkage_validate()` on classroom-budget linkage | 10 |
| `linkage_validate()` on student-classroom linkage | 11 |
| `linkage_validate()` on a master object | 13 |

Code that consumes validation results should use `nrow(validation$checks)`
rather than hard-coding a total count.

## Severity Levels

| Status | Meaning | Public interpretation |
|--------|---------|-----------------------|
| `PASS` | The check met its rule, or the rule is not applicable for this object. | No action required. |
| `WARN` | A potential quality issue or coverage caveat needs review. | Default mode may proceed; strict mode fails. |
| `ERROR` | A required contract was violated or a serious linkage problem was detected. | Do not proceed until repaired or explicitly resolved. |
| `INFO` | Context that should be reported but is not a failure. | Preserve in logs and summaries. |

`INFO` is especially important for administrative data. A missing module year
can be expected coverage metadata rather than a failed join.

## Synthetic Example With Coverage Gaps

This vignette uses synthetic data only. The example intentionally gives budget
and student panels fewer years than the classroom panel so the validation
framework has coverage information to report.

```{r build-synthetic-master}
budget <- alprek_synthetic_budget(
  n_classrooms = 12,
  n_years = 2,
  seed = 42
)

classroom <- alprek_synthetic_classroom(
  n_classrooms = 12,
  n_years = 3,
  seed = 42
)

student <- alprek_synthetic_student(
  n_students = 60,
  n_classrooms = 12,
  n_years = 2,
  seed = 42
)

student <- suppressMessages(student_transform(student))
master <- suppressMessages(linkage_create_master(budget, classroom, student))
validation <- suppressMessages(linkage_validate(master))
```

```{r validation-summary}
validation_summary <- function(x) {
  data.frame(
    passed = x$passed,
    n_errors = x$n_errors,
    n_warnings = x$n_warnings,
    n_info = x$n_info,
    n_checks = nrow(x$checks)
  )
}

validation_summary(validation)
```

The validation object reports the status counts. It does not require printing
student records, classroom rows, source paths, or private identifiers.

## Strict Mode

Default validation passes when there are no `ERROR` checks. Strict mode passes
only when there are no `ERROR` and no `WARN` checks. `INFO` checks never make an
object fail.

```{r strict-mode}
validation_default <- suppressMessages(linkage_validate(master, strict = FALSE))
validation_strict <- suppressMessages(linkage_validate(master, strict = TRUE))

data.frame(
  mode = c("default", "strict"),
  passed = c(validation_default$passed, validation_strict$passed),
  errors = c(validation_default$n_errors, validation_strict$n_errors),
  warnings = c(validation_default$n_warnings, validation_strict$n_warnings),
  info = c(validation_default$n_info, validation_strict$n_info),
  checks = c(nrow(validation_default$checks), nrow(validation_strict$checks))
)
```

Strict mode is useful when a workflow needs a clean handoff before downstream
analysis. Default mode is useful when a warning is expected and documented.

## Coverage-Aware Linkage Diagnostics

The master object stores coverage by school year.

```{r coverage-table}
master$diagnostics$coverage$by_year[, c(
  "school_year",
  "has_budget",
  "has_classroom",
  "has_student",
  "budget_status"
)]
```

The corresponding validation checks separate expected coverage gaps from true
overlap-year linkage problems.

```{r coverage-checks}
coverage_checks <- c(
  "budget_overlap_orphans",
  "budget_missing_coverage",
  "student_classroom_missing_coverage",
  "student_classroom_overlap_orphans",
  "empty_classrooms",
  "na_introduced",
  "year_coverage"
)

validation$checks[
  validation$checks$check_name %in% coverage_checks,
  c("check_name", "status", "n_issues", "details")
]
```

The key distinction is:

- missing budget coverage is explicit metadata;
- rows in missing budget years retain missing budget-derived fields;
- overlap-year classroom-budget or student-classroom orphans are real linkage
  problems;
- classroom years without student coverage are warnings because classroom rows
  are retained without linked student rows;
- student years without classroom coverage are errors because student rows lack
  the classroom context required for the linked master.

## Aggregate Transparency

Validation output should be summarized before it leaves a private workflow.
Status counts and issue totals are usually enough for public documentation.

```{r aggregate-status}
aggregate(
  n_issues ~ status,
  data = validation$checks,
  FUN = sum
)
```

Analysts can keep the full validation object in a private run log, but public
vignettes and reports should avoid row-level excerpts from confidential data.

## Hidden Contract Checks

The vignette itself checks the object contracts without printing private data.

```{r contract-checks, include = FALSE}
stopifnot(inherits(master, "alprek_linkage_master"))
stopifnot(all(c(
  "check_name",
  "check_description",
  "status",
  "n_issues",
  "details"
) %in% names(validation$checks)))
stopifnot(all(validation$checks$status %in% c("PASS", "WARN", "ERROR", "INFO")))
stopifnot(!any(grepl("[.]x$|[.]y$", names(master$classroom_level))))
stopifnot(!any(grepl("[.]x$|[.]y$", names(master$student_level))))
stopifnot(validation_default$passed)
stopifnot(!validation_strict$passed)
```

## What Public Vignettes Should Not Print

Public validation examples should not print:

- source-data file paths;
- row-level private records;
- student identifiers;
- names, dates of birth, or staff contact fields;
- raw classroom codes from confidential files;
- full private exports;
- real ADECE counts unless they have already been cleared for publication.

Synthetic values are useful for testing the pipeline, but they are not empirical
estimates from the real program.

## Reading Map

| Need | Where to go |
|------|-------------|
| Linkage workflow examples | A3 |
| DuckDB round-trip validation | A4 |
| Targets workflow validation summaries | A5 |
| Codebook and mapping contracts | M3 |
| Privacy and provenance policy | M4 |
