---
title: "M3 - Codebooks and Mappings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{M3 - Codebooks and Mappings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

ALprekDB keeps many data-cleaning decisions outside R code. Public CSV files in
`inst/extdata/` define column mappings and codebooks that are read by the
processing functions at runtime.

This design has three practical benefits:

- codebooks are readable by analysts and program staff who do not write R;
- mapping changes are versioned as data-cleaning decisions;
- public documentation can describe processing rules without sharing raw ADECE
  records.

```{r load}
library(ALprekDB)
```

## Installed CSV Inventory

The package ships 29 public CSV files in v0.8.0: 19 codebooks and 10 column
mappings. Eight of those files support the Applications module (introduced in
v0.7.0) and seven support the Geocode module (introduced in v0.8.0).

```{r inventory}
ext_dir <- system.file("extdata", package = "ALprekDB", mustWork = TRUE)
ext_files <- list.files(
  ext_dir,
  pattern = "[.]csv$",
  recursive = TRUE,
  full.names = TRUE
)
ext_relative_files <- sub("^.*[/]extdata[/]", "", ext_files)

csv_inventory <- data.frame(
  file = ext_relative_files,
  role = ifelse(grepl("^mappings/", ext_relative_files), "mapping", "codebook"),
  rows = unname(vapply(
    ext_files,
    function(path) nrow(utils::read.csv(path, check.names = FALSE)),
    integer(1)
  )),
  fields = unname(vapply(
    ext_files,
    function(path) length(names(utils::read.csv(path, nrows = 0, check.names = FALSE))),
    integer(1)
  )),
  row.names = NULL
)

csv_inventory[order(csv_inventory$role, csv_inventory$file), ]
```

```{r inventory-summary}
aggregate(
  cbind(files = rep(1L, nrow(csv_inventory)), rows = csv_inventory$rows),
  by = list(role = csv_inventory$role),
  FUN = sum
)
```

These files are part of the package API in a broad sense: changing them changes
how raw fields become standardized analytic columns.

## Two Kinds of CSV Files

Column mappings have a common four-column schema.

```{r mapping-schema}
mapping_files <- csv_inventory[csv_inventory$role == "mapping", "file"]

data.frame(
  file = mapping_files,
  schema = vapply(
    file.path(ext_dir, mapping_files),
    function(path) {
      paste(names(utils::read.csv(path, nrows = 0, check.names = FALSE)), collapse = ", ")
    },
    character(1)
  ),
  row.names = NULL
)
```

| Mapping field | Meaning |
|---------------|---------|
| `raw_column` | Column label expected in a source file format. |
| `standard_name` | Package-standard variable name used after cleaning. |
| `type` | Domain tag used for cleaning, privacy filtering, and tests. |
| `notes` | Human-readable explanation or implementation note. |

Codebooks have domain-specific schemas because they serve different cleaning
tasks.

```{r helper-schemas}
helpers <- list(
  delivery_types = alprek_delivery_types(),
  county_codes = alprek_county_codes(),
  degree_patterns = alprek_degree_patterns(),
  budget_categories = alprek_category_groups(),
  classroom_race = alprek_race_mapping(),
  classroom_language = alprek_language_mapping(),
  student_race = alprek_student_race_mapping(),
  student_delivery = alprek_student_delivery_mapping(),
  applications_status = alprek_applications_status_codes(),
  applications_funding = alprek_applications_funding_types(),
  applications_manifest = alprek_applications_source_manifest(),
  applications_edge_cases = utils::read.csv(
    file.path(ext_dir, "codebooks", "applications_edge_cases.csv"),
    stringsAsFactors = FALSE,
    check.names = FALSE
  ),
  geocode_column_map = alprek_geocode_column_map(),
  geocode_resultcode = alprek_geocode_resultcode_meaning(),
  geocode_statuscode = alprek_geocode_statuscode_meaning(),
  geocode_errorcode = alprek_geocode_errorcode_meaning(),
  geocode_al_fips = alprek_geocode_al_fips_counties(),
  geocode_manifest = alprek_geocode_source_manifest(),
  geocode_edge_cases = utils::read.csv(
    file.path(ext_dir, "codebooks", "geocode_edge_cases.csv"),
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
)

data.frame(
  helper = names(helpers),
  rows = vapply(helpers, nrow, integer(1)),
  schema = vapply(helpers, function(x) paste(names(x), collapse = ", "), character(1)),
  row.names = NULL
)
```

## Codebooks as Boundary Objects

A codebook can be read in a spreadsheet, reviewed in a pull request, and loaded
by the package. The delivery-type codebook is a simple example.

```{r delivery-types}
alprek_delivery_types()
```

The same seven provider types are used by classroom codes, classroom panels,
student panels, linkage diagnostics, and synthetic data.

The degree-pattern codebook is more procedural. It defines regular-expression
patterns and priorities for parsing free-text credential fields into standard
degree levels and areas. The codebook is still data, but it drives executable
cleaning logic.

```{r degree-pattern-types}
aggregate(
  priority ~ pattern_type,
  data = alprek_degree_patterns(),
  FUN = length
)
```

## How Mappings Are Used

The module cleaning functions load mappings after format detection.

```text
read source columns
  -> detect format
  -> load matching column map
  -> rename raw columns to standard names
  -> apply codebooks and derived variables
  -> validate standardized output
```

For example:

- budget cleaning uses `budget_category_groups.csv` to group detailed budget
  line items;
- classroom cleaning uses classroom mappings, degree patterns, race mappings,
  language mappings, and delivery-type codes;
- student cleaning uses student mappings, student race mappings, and student
  delivery-type mappings;
- applications cleaning uses four cycle-1 column maps plus status-code,
  funding-type, source-manifest, and edge-case codebooks;
- geocode cleaning uses the Melissa column-map codebook together with
  RESULTCODE / STATUSCODE / ERRORCODE meaning codebooks, the Alabama FIPS
  county codebook, the geocode source manifest, and a geocode edge-case
  codebook.

The public mappings describe how raw field names are standardized; they should
not contain row-level examples from real source data.

## Format Detection

Column mappings are selected by detected format. Detection rules look at column
names, marker fields, and source-file width rather than row-level values.

```{r format-detection-demo}
budget_legacy <- data.frame(
  "Classroom Code" = "",
  "Classroom Name" = "",
  "Example From OSR Funds" = 0,
  check.names = FALSE
)

budget_new <- data.frame(
  "Classroom Code" = "",
  "Classroom Name" = "",
  "OSR" = 0,
  "Proration" = 0,
  check.names = FALSE
)

classroom_legacy <- as.data.frame(matrix(NA, nrow = 1, ncol = 100))
names(classroom_legacy) <- paste0("col", seq_len(ncol(classroom_legacy)))

classroom_new <- as.data.frame(matrix(NA, nrow = 1, ncol = 121))
names(classroom_new) <- paste0("col", seq_len(ncol(classroom_new)))

student_legacy <- as.data.frame(matrix(NA, nrow = 1, ncol = 202))
names(student_legacy) <- paste0("col", seq_len(ncol(student_legacy)))

student_new <- as.data.frame(matrix(NA, nrow = 1, ncol = 250))
names(student_new) <- paste0("col", seq_len(ncol(student_new)))

data.frame(
  module = c("budget", "budget", "classroom", "classroom", "student", "student"),
  scenario = c("legacy marker", "new marker", "legacy width", "new width", "legacy width", "new width"),
  detected = c(
    budget_detect_format(budget_legacy),
    budget_detect_format(budget_new),
    classroom_detect_format(classroom_legacy),
    classroom_detect_format(classroom_new),
    student_detect_format(student_legacy),
    student_detect_format(student_new)
  )
)
```

Format detection is deliberately conservative. If a source export is too narrow
or contains incompatible markers, the package stops rather than guessing.

## Applications Codebooks and Mappings

The v0.7.0 Applications module adds four public codebooks and four cycle-1
column maps:

```{r applications-extdata}
csv_inventory[
  grepl("^codebooks/applications_|^mappings/applications_", csv_inventory$file),
  c("file", "role", "rows", "fields")
]
```

The non-renewal source sheet in cycle-1 is headerless, so
`applications_column_map_nonrenewals_cycle1.csv` maps positional `col_1`,
`col_2`, ... fields. Future cycles should add new cycle-specific mapping files
rather than editing old maps in place.

## Geocode Codebooks

The v0.8.0 Geocode module adds seven public codebooks. All seven live in the
`codebooks/` directory (the Melissa column map is stored as a codebook, not a
mapping, because it has a Melissa-specific schema rather than the standard
four-column mapping contract).

```{r geocode-extdata}
csv_inventory[
  grepl("^codebooks/(geocode_|melissa_)", csv_inventory$file),
  c("file", "role", "rows", "fields")
]
```

The RESULTCODE / STATUSCODE / ERRORCODE codebooks drive validation. Rather
than hard-coding valid codes, `geocode_validate()` reads the codebooks and
flags any row whose parsed codes are not present in the corresponding
codebook. This keeps validation in step with Melissa documentation changes
without R-code edits.

## Mapping Contracts

The tests enforce these contracts:

- all 29 expected CSV files are present;
- mapping files use `raw_column`, `standard_name`, `type`, and `notes`;
- mapping keys are nonblank and unique within each file;
- `standard_name` values are unique and use snake case;
- codebook keys are unique;
- canonical domains such as delivery types and race categories are covered;
- degree-pattern regexes compile;
- privacy-related mapping types remain explicit in new-format maps.

```{r mapping-contracts}
all_mappings <- lapply(file.path(ext_dir, mapping_files), function(path) {
  utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
})
names(all_mappings) <- mapping_files

data.frame(
  file = names(all_mappings),
  unique_raw_column = vapply(all_mappings, function(x) !any(duplicated(x$raw_column)), logical(1)),
  unique_standard_name = vapply(all_mappings, function(x) !any(duplicated(x$standard_name)), logical(1)),
  snake_case_standard_name = vapply(
    all_mappings,
    function(x) all(grepl("^[a-z][a-z0-9_]*$", x$standard_name)),
    logical(1)
  ),
  row.names = NULL
)
```

## Extension Workflow

When a source format changes, the safest update path is:

1. Add or update a mapping CSV with the same schema.
2. Add or update a detection rule that selects the mapping only for the intended
   format.
3. Keep raw data and row-level examples out of `inst/extdata/`.
4. Add schema tests for the mapping file and its required standardized names.
5. Add cleaning tests that exercise the new mapping on synthetic or minimal
   fixture data.
6. Run the affected module tests and render the relevant vignettes.

This keeps format drift visible as a versioned package change rather than a
private one-off script edit.

## What Public Codebooks Should Not Contain

Do not put these in public package codebooks or mappings:

- raw ADECE files or row extracts;
- real classroom codes from confidential files;
- student identifiers;
- names, dates of birth, or staff contact fields;
- local absolute paths;
- private output filenames;
- examples copied from a private worksheet.

Public codebooks can describe categories, standardized names, and format-level
field mappings. They should not document individual records.

## Hidden Contract Checks

```{r contract-checks, include = FALSE}
stopifnot(nrow(csv_inventory) == 29L)
stopifnot(sum(csv_inventory$role == "codebook") == 19L)
stopifnot(sum(csv_inventory$role == "mapping") == 10L)
stopifnot(sum(csv_inventory$rows[csv_inventory$role == "mapping"]) == 755L)
stopifnot(all(c("code", "name", "name_short") %in% names(alprek_delivery_types())))
stopifnot(setequal(alprek_delivery_types()$code, c("P", "C", "H", "O", "F", "U", "S")))
stopifnot(all(vapply(all_mappings, function(x) {
  all(c("raw_column", "standard_name", "type", "notes") %in% names(x))
}, logical(1))))
```

## Reading Map

| Need | Where to go |
|------|-------------|
| Why externalized codebooks are part of the architecture | M1 |
| How validation checks enforce codebook contracts | M2 |
| How to build panels with current mappings | A2 |
| How to keep private source files out of public docs | M4 |
